---
title: 'Tutorial 1: Introduction to Machine Learning with Python¶'
author: "Jonathan Ish-Horowicz"
date: "12/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tutorial 1: Introduction to Machine Learning with Python

The goal of this tutorial is to introduce a typical workflow in carrying out ML in R.  Similarly to last week's Python tutorial, this includes:

1. accessing and organising data,
2. assessing the data,
3. visualising the data,
4.  a) creating training, b) test datasets and c) learning a model using them and evaluating its performance.


# 1) Load Data

We load the same iris dataset as last week, which has 150 samples and 4 attributes. There are 3 classes (species).

In R we can load the Iris dataset from the `datasets` package:

```{r message=FALSE}
# Load the iris dataset
library(datasets)
data(iris)
iris$Species <- as.character(iris$Species)
```

# 2) Statistics of the dataset

Now compute the mean, standard deviation, minimum and maximum of each attribute.

Suggestion: use the the `group_by` and `summarise_all` functions from the `dplyr` library.

```{r message=FALSE}
library(dplyr)

# Calculate the mean of each attribute
iris_means <- iris %>%
  group_by(Species) %>%
  summarise_all(mean)
iris_means

# Calculate the standard deviation of each attribute
iris_sd <- iris %>%
  group_by(Species) %>%
  summarise_all(sd)
iris_sd

# Calculate the minimum of each attribute
iris_all <- iris %>%
  group_by(Species) %>%
  summarise_all(list(min = min, max = max, mean = mean, stdev = sd))
iris_all
# Calculate the maximum of each attribute

```

# 3) Visualise the dataset

Make some exploratory plots here.

Suggestion: use the `ggplot2` library. For a nice pairs plot use the `ggpairs` function from `GGally` (a `ggplot2` extension library).

```{r message=FALSE}
library(ggplot2)
library(GGally)

corr <- cor(iris[,1:4])
corr
plot <- ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point(aes(col = Species))
plot
plot2 <- ggplot(iris, aes(x = Petal.Width, y = Sepal.Width)) + 
  geom_point(aes(col = Species))
plot2
plot3 <- ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point(aes(col = Species))
plot3
plot4 <- ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) + 
  geom_point(aes(col = Species))
plot4

ggpairs(iris, aes(color = Species, alpha = 0.5), progress = FALSE)
```

# 4) Classification using Least Squares

Here we will be carrying out classification using the least squares formulation on 2 classes of the dataset.

a) Create separate datasets for the classes `setosa` and `versicolor`.

```{r}
# your code here
setosa <- iris %>%
  filter(Species == "setosa")
versicolor <- iris %>%
  filter(Species== "versicolor")
# result should be two dataframes (one for each for setosa, versicolor classes),
# each with dim (50,5) - 4 attributes plus column for class
stopifnot(all(dim(setosa)==c(50,5)))
stopifnot(all(dim(versicolor)==c(50,5)))
```


b) add a column to each dataset where the column is $1$ if the class is `setosa` and $-1$ otherwise.

```{r}
# your code here
setosa$output <- rep(1, length(setosa$Species))
versicolor$output <- rep(-1, length(versicolor$Species))
# result should add a column to each of setosa and versicolor
stopifnot(all(dim(setosa)==c(50,6)))
stopifnot(all(dim(versicolor)==c(50,6)))
```


c) create training and test datasets, with 20% of the data for testing. This 80 training points and 20 testing points in total (half this per class).

```{r}
# your code here - see improved version!!
set.seed(311083)
rows <- sample(nrow(setosa))
setosa <- setosa[rows,]
versicolor <- versicolor[rows,]
training.data <- rbind(setosa[1:40,], versicolor[1:40,])
test.data <- rbind(setosa[41:50,], versicolor[41:50,])

# resulting dataframes (one each for training and test data) should have
# the appropriate sizes
stopifnot(all(dim(training.data)==c(80,6)))
stopifnot(all(dim(test.data)==c(20,6)))
```



d) apply the least squares solution to obtain an optimal solution for different combinations of the 4 available attributes. The code to create a list containing all the combinations of the attributes has been provided.

```{r}
# Creates all possible combinations of attributes
# attribute.combinations is a list whose elements are lists of attributes
attribute.names <- colnames(iris)[1:4]
attribute.combinations <- do.call(
  c,
  lapply(1:4, function(i) as.list(data.frame(combn(attribute.names, i))))
  )
names(attribute.combinations) <- 1:length(attribute.combinations)
attribute.names

return.predictions <- function(attribute.names, training.data, test.data) {
  
  # Format training and test data (as matrices)
  X_train = as.matrix(training.data[as.character(attribute.names)])
  y_train = as.vector(training.data$output)
  X_test = as.matrix(test.data[as.character(attribute.names)])
  y_test = as.matrix(test.data$output)

  # Calculate optimal weights
  w = solve(t(X_train)%*%X_train)%*%t(X_train)%*%y_train
  
  # Make predictions
  predictions = X_test%*%w

  return(predictions)
}

return.predictions(attribute.names, training.data, test.data)
p1 <- return.predictions(attribute.combinations[[1]], training.data, test.data)
p2 <- return.predictions(attribute.combinations[[2]], training.data, test.data)
```



e) evaluate which input attributes are the best.

```{r}
predictions <- return.predictions(attribute.names, training.data, test.data)
length(predictions)
# Calculate the mean square error between some predictions and
# the corresponding testing data
return.mse <- function(predictions, test.data) {
  y.test <- as.numeric(test.data$class)
  error <- y.test - predictions
  square.error <- error**2.0
  mse <- mean(square.error)
  return(mse)
}
return.mse(predictions, test.data)
# Calculate the test MSE for each the elements of attribute.combinations
# by calling return.predictions and return.mse
mse.df <- as.data.frame(
  sapply(attribute.combinations, 
         function(x) return.mse(
           return.predictions(x, training.data, test.data),
           test.data)
         )
  )
colnames(mse.df) <- c("MSE")
mse.df$attributes <- sapply(attribute.combinations, function(x) paste(x, collapse = ","))

paste(attribute.combinations[[10]], collapse = ",") #to understand
ggplot(mse.df, aes(x = attributes, y = MSE )) + 
  geom_point()
##sapply: Arg1: List (input will be elements of the list), 
##Arg2: Function with one argument/operation
```